---
title: "More Work"
date: "`r Sys.Date()`"
author:  Homer White
output:
  rmdformats::downcute:
    toc_depth: 2
    self_contained: true
    highlight:  tango
    lightbox: true
    code_download: true
---


```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(tigerstats)
## Global options
options(dplyr.inform.summarize = FALSE)
knitr::opts_chunk$set(
  out.width = "90%",
  fig.align = "center",
  tidy = FALSE,
  warning = FALSE,
  message = FALSE
)
```



## Import the Data

```{r}
df <- read_xlsx(
  "data/Blastexposure.xlsx"
)
```

Summarize the data set to get an idea of the variables and their values:

```{r}
summary(df)
```

That didn't tell me much.  I should figure out what's in `DiagA`:

```{r}
df %>% 
  pull(DiagA) %>% 
  unique() %>% 
  sort()
```

Seems clear enough!

## Some Tables

OK, let's see if there is a relationship between exposure to blasts and PTSD:

```{r}
tab2 <- xtabs(~ NumBlasts + DiagA, data = df)
tab2
```

Holy moly!  Looks like the more blasts you get, the greater the likelihood of PTSD.

Let's rename and recode a bit:

```{r}
df2 <- 
  df %>% 
  rename(diagnosis = DiagA) %>% 
  mutate(blast_exposure = case_when(
    NumBlasts == 0 ~ "0 blasts",
    NumBlasts == 1 ~ "1 blast",
    NumBlasts == 2 ~ "2+ blasts"
    )) %>% 
  mutate(blast_exposure = factor(
    blast_exposure,
    levels = c("0 blasts", "1 blast", "2+ blasts")
  ))
```


Another table:

```{r}
df2 %>%
  drop_na(blast_exposure, diagnosis) %>% 
  group_by(blast_exposure) %>% 
  summarize(
    PTSD = sum(diagnosis == "PTSD"),
    total = n(),
    percent = PTSD / total
  )
```

## Some Graphs

Go for a bar graphs:

```{r fig.cap = "nice caption"}
df2 %>% 
  xtabs(~ blast_exposure + diagnosis, data = .) %>% 
  mosaicplot(
    main = "Mosic Plot of Diagnosi vs. Blast Exposure",
    xlab = "Number of blasts",
    ylab = "Diagnosis"
  )
```

## Some Inferential Statistics

A $\chi^2$-test is appropriate here:

```{r}
chisqtestGC(
  ~ blast_exposure + diagnosis,
  data = df2
)
```

We have overwhelmingly strong evidence that folks exposed to more blasts are more likely to be diagnosed with PTSD.



